Transparency of COVID-19-related research in medical journals

### 2022-04-23 manuscript_covid_transparency.qmd
changes in the intro and new references

COVID-19 pandemic launched an incredibly fast and voluminous avalanche of scientific publications across the different fields of science. There has been discussion and many examples related to poor or suboptimal data, methods and other scientific practices in COVID-19-related studies \[@Glasziou2020; @besançon2021; @luengo-oroz2020; @schwab2020\]. However, in addition to fast and successful development of safe and effective vaccination, it is possible that transparent scientific practices, like data and code sharing, have helped to fight pandemic globally \[@besançon2021; @vuong2022\]. That being said, comprehensive picture of transparency practices in COVID-19-related medical literature is still lacking. As an example, it is unknown whether open science initiatives related to COVID-19 pandemic \[@besançon2021; @vuong2022\], for instance to advance data sharing, has led to higher number of articles sharing data in COVID-19 research compared to what has been seen in studies in general in biomedical literature [@serghiou2021].

@@

Adherence to transparent practices was at similar level in COVID-19 related literature than in other biomedical literature analyzed with the same methods (see Fig 3 in [@serghiou2021\]). This is surprising particularly when considering worldwide, remarkable and noble initiatives related to enhancing open science to tackle the pandemic. As early as in January 2020, over 100 organizations, including journals, publishers, funders, universities and other institutions, signed a statement that aimed to ensure free access to research data, tools and other information related to COVID-19 \[@vuong2022; @coronavi]. Later also other initiatives to support the goal emerged, for instance COVID-19 Open Research Dataset (CORD-19)[@cord-19], a free resource of over 280,000 articles about the COVID-19 virus. However, it is possible that the algorithms used here did not efficiently detect all different ways of sharing data and material emerged after the pandemic, because the algorithms were validated before the pandemic [@serghiou2021].

In comparison to other studies on the transparency of COVID-19 related research is difficult due to different methodologies and scale of our research. However, it seems that data sharing was little less common than in our sample than what was the proportion detected in the beginning of the pandemic measured by the investigation of associated data in PubMed Central [@lucas-dominguez2021\]. In addition, the proportion of studies which shared data in our study sample was lower than in COVID-19 related preprints shared via medRxiv and bioRxiv \[@strcic2022\]. Lacking protocol registration, sharing code and COI have got attention during the pandemic \[@Glasziou2020; @besançon2021; @luengo-oroz2020], however we are not aware of the investigation of transparency in these aspects in COVID-19 related research in this scale in medical research.

>>>>

During the recent COVID-19 pandemic, part of the public in developed countries justified their lack of adherence to public health measures due to the lack of transparency of the scientific studies used as an argument by national authorities [@Skipper2020-rs\]. Thus, supranational bodies such as the OECD indicate that transparent communication to citizens is vital to support public health measures and to counter misinformation \[@Bellantoni_undated-tf\]. On the other hand, the COVID-19 pandemic launched an incredibly fast and voluminous avalanche of scientific publications across the different fields of science. This research led to the development of vaccines as well as the evaluation of community interventions. On the one hand, it is possible that transparent scientific practices, like data and code sharing, have helped to fight pandemic globally \[@besançon2021; @vuong2022\], while on the other hand, the lack of transparency of some key developments has led the public to mistrust the results of research and public health measures \[@Skipper2020-rs\]. This shows that there is a lack of awareness about the level of transparency practices in COVID-19-related medical literature. As an example, it is unknown whether open science initiatives related to COVID-19 pandemic \[@besançon2021; @vuong2022] to advance data sharing, has led to higher number of articles sharing data in COVID-19 research compared to what has been seen in studies in general in biomedical literature [@serghiou2021].

@@

Adherence to transparent practices was at similar level in COVID-19 related literature than in other biomedical literature analyzed with the same methods (see Fig 3 in \[@serghiou2021\]). This is surprising particularly when considering worldwide, remarkable and noble initiatives related to enhancing open science to tackle the pandemic. As early as in January 2020, over 100 organizations, including journals, publishers, funders, universities and other institutions, signed a statement that aimed to ensure free access to research data, tools and other information related to COVID-19 \[@vuong2022; @coronavi\]. Later also other initiatives to support the goal emerged, for instance COVID-19 Open Research Dataset (CORD-19)[@cord-19], a free resource of over 280,000 articles about the COVID-19 virus. However, it is possible that the algorithms used here did not efficiently detect all different ways of sharing data and material emerged after the pandemic, because the algorithms were validated before the pandemic [@serghiou2021].

In comparison to other studies on the transparency of COVID-19 related research is difficult due to different methodologies and scale of our research. However, it seems that data sharing was little less common than in our sample than what was the proportion detected in the beginning of the pandemic measured by the investigation of associated data in PubMed Central \[@lucas-dominguez2021\]. In addition, the proportion of studies which shared data in our study sample was lower than in COVID-19 related preprints shared via medRxiv and bioRxiv \[@strcic2022\]. Lacking protocol registration, sharing code and COI have got attention during the pandemic \[@Glasziou2020; @besançon2021; @luengo-oroz2020\], however we are not aware of the investigation of transparency in these aspects in COVID-19 related research in this scale in medical research.
